%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Lecture notes for Numerical Methods for Partial Differential Equations
%
% Chapter 2: Parabolic PDEs
%   Sections 3 and 4
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% !TeX root = NumPDE_Lecture_notes.tex

\subsection{Implicit methods}

\subsubsection{Backward-difference method}
 
To obtain a method which is {\bf unconditionally stable}, we consider
an implicit-difference method that results from using the backward-difference formula for
$\pr u/\pr t(x,t)$:
\[
\frac{\pr u}{\pr t}(x_{k},t_{j})=\frac{u(x_{k},t_{j})-u(x_{k},t_{j-1})}{\tau}+
\frac{\tau}{2}\frac{\pr^{2} u}{\pr t^{2}}(x_{k}, \mu_{j}),
\]
where $\mu_{j}\in(t_{j-1},t_{j})$. Using this instead of the forward-difference formula
yields the following difference equation
\begin{equation}
\frac{w_{kj}-w_{k,j-1}}{\tau}-K
\frac{w_{k+1, j}-2w_{kj}+w_{k-1,j}}{h^{2}}=0 \label{x9}
\end{equation}
for $k=1, 2, \dots, N-1$ and $j=1, 2, \dots,M$. This equation is different from Eq. (\ref{b4})
and cannot be solved using a procedure as straightforward as that
in the forward-difference method.

 
 
Equation (\ref{x9}) can be rewritten as
\begin{equation}
(1+2\gamma)w_{kj}-\gamma (w_{k+1, j}+w_{k-1, j})=w_{k,j-1} \label{x10}
\end{equation}
for $k=1, 2, \dots, N-1$ and $j=1, 2, \dots,M$, or, equivalently,
\begin{equation}
A{\bf w}^{(j)}={\bf w}^{(j-1)} \quad \hbox{for} \quad j=1,2,\dots, \label{x11}
\end{equation}
where
\begin{equation}
A=\left[
\begin{array}{cccccc}
1+2\gamma &-\gamma &0      &\dots  &\dots &0 \\
-\gamma &1+2\gamma &-\gamma &\ddots  &     &\vdots \\
0      &-\gamma &1+2\gamma &-\gamma &\ddots &\vdots \\
\vdots &\ddots &\ddots &\ddots &\ddots &0 \\
\vdots &       &\ddots &\ddots &\ddots &-\gamma \\
0      &\dots  &\dots  &0      &-\gamma &1+2\gamma
\end{array}\right], \quad
{\bf w}^{(j)}=\left[
\begin{array}{c}
w_{1,j} \\
w_{2,j} \\
\vdots \\
\vdots \\
\vdots \\
w_{N-1,j}
\end{array}\right].
\label{x12}
\end{equation}
If we know ${\bf w}^{(j-1)}$, Eq. (\ref{x11}) represents a system of linear equations for
${\bf w}^{(j)}$ which can be solved, e.g., by the double-sweep method that we will describe in 
the next subsection.

The finite-difference method described above is called the {\bf backward-difference method}. It is an example
of {\bf implicit} finite-difference methods.

 
 
Now let us investigate the stability of the backward-difference scheme
by the Fourier method.
If we introduce the error $z_{k0}=w_{k0}-u_{0}(x_{k})$ into the initial condition, it
will propagate with each step in time. Let $z_{kj}=w_{kj}-u_{0}(x_{k},t_{j})$ be
the error at the mesh point $(x_{k}, t_{j})$ for each $k=0,1,2, \dots, N$
and $j=0,1, \dots$.
It follows from (\ref{x9}) that $z_{kj}$ satisfies the difference equation
\begin{equation}
\frac{z_{kj}-z_{k,j-1}}{\tau}-K
\frac{z_{k+1, j}-2z_{kj}+z_{k-1,j}}{h^{2}}=0  \label{c15}
\end{equation}
for $k=1, 2, \dots, N-1$ and $j=1, 2, \dots, M$.
We seek a particular solution of
(\ref{c15}) in the form
\begin{equation}
z_{k,j}=\rho_{q}^{j}e^{iqx_{k}}, \quad q\in{\mathbb R}. \label{c16}
\end{equation}
The finite-difference method (\ref{x9}) is stable, if
\[
\vert\rho_{q}\vert\leq 1
\]
for all $q$.

 
 
Substitution of (\ref{c16}) into (\ref{c15}) yields
\[
e^{iqx_{k}}\left(\rho_{q}^{j}-\rho^{j-1}\right)-
\gamma\rho_{q}^{j}\left(e^{iqx_{k+1}}-2e^{iqx_{k}}+e^{iqx_{k-1}}\right)=0
\]
or
\[
1-\frac{1}{\rho_{q}}-
\gamma\left(e^{iqh}-2+e^{-iqh}\right)=0.
\]
Since
\[
e^{iqh}-2+e^{-iqh}=\left(e^{iqh/2}-e^{-iqh/2}\right)^{2}=-4\sin^{2} \frac{qh}{2},
\]
we obtain
\[
\rho_{q}=\frac{1}{1+4\gamma\sin^{2} \frac{qh}{2}}.
\]
Evidently, $\vert\rho_{q}\vert\leq 1$ for all $q$, and therefore, the method is stable.
Note that the method is {\bf unconditionally stable} (independent
of the choice of $\gamma=K\tau/h^{2}$). It can be shown that
its local truncation
error is $O(\tau+h^{2})$ (the same as in the forward-difference method).

\subsubsection{Double-sweep algorithm}
 
\textbf{The double-sweep method.} How to solve system (\ref{x10})? This system is tridiagonal. The most efficient algorithm
of solving such systems is called {\bf the double-sweep method}. Consider the following
tridiagonal system:
\begin{equation}
A_{i}v_{i-1}-C_{i}v_{i}+B_{i}v_{i+1}=F_{i} \quad \hbox{for}\quad i=1, \dots, N-1;
\quad v_{0}=v_{N}=0; \label{y22}
\end{equation}
where the coefficients $A_{i}$, $B_{i}$ and $C_{i}$ satisfy the conditions
\begin{equation}
A_{i}, B_{i}, C_{i} > 0, \quad  C_{i} \geq A_{i} + B_{i}.  \label{y23}
\end{equation}
(One can verify that system (\ref{x10}) satisfies the conditions (\ref{y23}).)
To solve (\ref{y22}), we will seek $\alpha_{i}$ and $\beta_{i}$ such that
\begin{equation}
v_{i-1}=\alpha_{i}v_{i}+\beta_{i}  \quad  \hbox{for} \quad
i=1, 2, \dots, N.  \label{y24}
\end{equation}
Substitution of (\ref{y24}) into (\ref{y22}) yields
\begin{equation}
(\alpha_{i}A_{i}-C_{i})v_{i}+B_{i}v_{i+1}+\beta_{i}A_{i}-F_{i}=0 \quad \hbox{for}\quad i=1, \dots, N-1. \label{y25}
\end{equation}
From (\ref{y24}), we also have
\[
v_{i}=\alpha_{i+1}v_{i+1}+\beta_{i+1}  \quad  \hbox{for} \quad
i=0, 1, \dots, N-1.
\]
Substituting this into (\ref{y25}), we find that
\[
[(\alpha_{i}A_{i}-C_{i})\alpha_{i+1}+B_{i}]v_{i+1}+[
(\alpha_{i}A_{i}-C_{i})\beta_{i+1}+\beta_{i}A_{i}-F_{i}]=0
\]
for $i=1, \dots, N-1$.
This equation is satisfied if the two expressions in the square brackets are both zero. 
This leads to the following recursive formulas:
\begin{equation}
\alpha_{i+1}=\frac{B_{i}}{C_{i}-\alpha_{i}A_{i}}, \quad
\beta_{i+1}=\frac{\beta_{i}A_{i}-F_{i}}{C_{i}-\alpha_{i}A_{i}}, \quad
\hbox{for}\quad i=1, \dots, N-1. \label{y26}
\end{equation}
Now if $\alpha_{1}$ and  $\beta_{1}$ are known, then $\alpha_{i}$ and  $\beta_{i}$ for $i=2, 3, \dots, N$
can be computed from Eqs. (\ref{y26}).  $\alpha_{1}$ and  $\beta_{1}$ can be determined from Eq. (\ref{y24}) and
the fact that $v_{0}=0$. Indeed
\[
v_{0}= \alpha_{1}v_{1}+\beta_{1} \quad \hbox{and} \quad v_{0}=0 \quad \Rightarrow \quad
\alpha_{1}v_{1}+\beta_{1}=0.
\]
To satisfy the last equation, we choose $\alpha_{1}=0$ and $\beta_{1}=0$. Once we know all $\alpha_{i}$
and $\beta_{i}$, we compute $v_{N-1}, v_{N-2}, \dots, v_{1}$ using formula (\ref{y24}).

   
Now let's show that conditions (\ref{y23}) are sufficient for the double-sweep method to work. Numbers $\alpha_{i}$ and $\beta_i$ are well-defined
by Eqs. (\ref{y26}) provided that $C_{i}-\alpha_{i}A_{i}\neq 0$ for $i=1,\dots,N$.
First we show that $0\leq\alpha_i < 1$ for $i=1,\dots,N$. We will show this by induction. Since $\alpha_1=0$, it satisfies
the required inequality. Now we need to show that if $0\leq\alpha_{i}<1$ for some
$i\geq 1$, then $0\leq\alpha_{i+1}<1$. it follows from (\ref{y23}) that
\[
C_{i}-\alpha_{i}A_{i}> A_i + B_i- \alpha_{i}A_{i} > B_i \quad
\Rightarrow \quad \alpha_{i+1}=\frac{B_{i}}{C_{i}-\alpha_{i}A_{i}} < 1.
\]
So, all $\alpha_i$ satisfy $0\leq\alpha_i < 1$. Therefore,
\[
C_{i}-\alpha_{i}A_{i}> A_i + B_i- \alpha_{i}A_{i} > B_i >0 \quad \hbox{for} \ \ i=1,\dots,N,
\]
as required.


\vskip 0.5cm  
\emph{Is it possible to generalise the double-sweep method to the case
when we have non-zero boundary conditions?} The answer is `Yes', and it is not difficult.

Suppose that we need to solve the same equations (\ref{y22}), but with non-zero boundary conditions:
\[
v_{0}=A, \quad v_{N}=B.
\]
The only thing that we need to modify is the choice of $\alpha_{1}$ and $\beta_{1}$. It follows from Eq. (\ref{y24})
and the boundary condition that
\[
\alpha_{1}v_{1}+\beta_{1}=A.
\]
To satisfy this we simply choose
\[
\alpha_{1}=0, \quad \beta_{1}=A.
\]
Boundary  condition $v_{N}=B$ is taken into account automatically when
$v_{N-1}$ is computed using Eq. (\ref{y24}).

\vskip 0.5cm  
The double-sweep method can also be generalised for
the following boundary conditions:}
\[
v_{0}=\lambda v_1+B \quad \hbox{and} \quad v_{N}=\mu v_{N-1} + B,
\]
where $\lambda$, $\mu$, $A$ and $B$ are constants and $\vert\lambda\vert\leq 1$, $\vert\mu\vert\leq 1$. (Prove this!)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Richardson method}

\emph{How to improve the order of approximation
of the heat equation by finite-difference schemes?} The easiest way
to obtain a finite-difference method whose truncation error is
$O(\tau^{2}+h^{2})$ is to replace the forward (backward)
difference formulas for $u_{t}$ with the central difference
formula:
\begin{equation}
\frac{\pr u}{\pr t}(x_k,t_j)=\frac{u(x_k,t_j+\tau)-u(x_k,t_j-\tau)}{2\tau}-
\frac{\tau^2}{6}\frac{\pr^3 u}{\pr t^3}(x_k,\xi) \label{aa15}
\end{equation}
[It is an exercise for you to derive this formula.]
This leads to the Richardson method, which for the heat equation gives
\begin{equation}
\frac{w_{k,j+1}-w_{k,j-1}}{2\tau}-K \frac{w_{k+1,
j}-2w_{kj}+w_{k-1,j}}{h^{2}}=0. \label{d4}
\end{equation}
(for $k=1, 2, \dots, N-1$ and $j=1, 2, \dots, M-1$). 
\footnote{This was the first published finite-difference methods for PDEs, introduced by Lewis Fry Richardson already in 1911 in
``The Approximate Arithmetical Solution by Finite Differences of Physical Problems Involving Differential Equations, with an Application to the Stresses in a Masonry Dam'', Philosophical Transactions of the Royal Society of London A 210, pp.307 -- 357.}


 
\begin{problem} Show that the Richardson method
has the local truncation error $O(\tau^{2}+h^{2})$.
\end{problem}
\begin{solution}
The local truncation error for Richardson's
method is
\begin{equation}
\tau_{kj}=\frac{u_{k,j+1}-u_{k,j-1}}{2\tau}-K \frac{u_{k+1,
j}-2u_{kj}+u_{k-1,j}}{h^{2}}. \label{d5}
\end{equation}
Here $u_{kj}=u(x_{k},t_{j})$. Expanding $u_{k,j\pm
1}=u(x_{k},t_{j\pm 1})=u(x_{k},t_{j}\pm \tau)$ in Taylor's series at
point $(x_{k},t_{j})$, we obtain
\begin{eqnarray}
&&u_{k,j+1}=u_{kj}+\tau \frac{\pr u}{\pr t}(x_k,t_j)+
\frac{\tau^2}{2}\frac{\pr^2 u}{\pr t^2}(x_k,t_j)+O(\tau^3),   \nonumber \\
&&u_{k,j-1}=u_{kj}-\tau \frac{\pr u}{\pr t}(x_k,t_j)+
\frac{\tau^2}{2}\frac{\pr^2 u}{\pr t^2}(x_k,t_j)+O(\tau^3).
\nonumber
\end{eqnarray}
It follows that
\begin{equation}
\frac{u_{k,j+1}-u_{k,j-1}}{2\tau}=\frac{\pr u}{\pr
t}(x_k,t_j)+O(\tau^2). \label{d6}
\end{equation}
Expanding $u_{k\pm 1,j}=u(x_{k\pm 1},t_j)=u(x_{k}\pm h,t_{j})$ in
Taylor's series at point $(x_{k},t_{j})$, we find that
\begin{eqnarray}
&&u_{k+1,j}=u_{kj}+h \frac{\pr u}{\pr x}(x_k,t_j)+
\frac{h^2}{2}\frac{\pr^2 u}{\pr x^2}(x_k,t_j)+
\frac{h^3}{6}\frac{\pr^3 u}{\pr x^3}(x_k,t_j)+O(h^4),   \nonumber \\
&&u_{k-1,j}=u_{kj}-h \frac{\pr u}{\pr x}(x_k,t_j)+
\frac{h^2}{2}\frac{\pr^2 u}{\pr x^2}(x_k,t_j)-
\frac{h^3}{6}\frac{\pr^3 u}{\pr x^3}(x_k,t_j)+O(h^4).   \nonumber
\end{eqnarray}
Hence,
\begin{equation}
\frac{u_{k+1, j}-2u_{kj}+u_{k-1,j}}{h^{2}}= \frac{\pr^2 u}{\pr
x^2}(x_k,t_j)+O(h^2). \label{d7}
\end{equation}
Substitution of (\ref{d6}) and (\ref{d7}) in Eq. (\ref{d5}) yields
\[
\tau_{kj}=\frac{\pr u}{\pr t}(x_k,t_j)-K \frac{\pr^2 u}{\pr
x^2}(x_k,t_j) +O(\tau^2)+O(h^2).
\]
Taking into account the fact that $u(x,t)$ satisfies the heat equation,
we obtain
\[
\tau_{kj}=O(\tau^2+h^2).
\]
\end{solution}
 
\begin{problem}
Investigate the
stability of Richardson's method, given by (\ref{d4}), by the
Fourier method.
\end{problem}
\begin{solution}
If $w_{kj}$ and $\tilde{w}_{kj}$ are two
solutions of the difference equation (\ref{d4}) that correspond to
different initial conditions, then the perturbation (error)
$z_{kj}=\tilde{w}_{kj}-w_{kj}$ satisfies the homogeneous difference
equation
\begin{equation}
\frac{z_{k,j+1}-z_{k,j-1}}{2\tau}-K \frac{z_{k+1,
j}-2z_{kj}+z_{k-1,j}}{h^{2}}=0. \label{d8}
\end{equation}
We seek a particular solution of (\ref{d8}) in the form
\[
z_{k,j}=\rho_{q}^{j}e^{iqx_{k}}.
\]
Substituting this into (\ref{d8}), we obtain
\[
e^{iqx_{k}}\left(\rho_{q}^{j+1}-\rho_{q}^{j-1}\right)-
2\gamma\rho_{q}^{j}
\left(e^{iqx_{k+1}}-2e^{iqx_{k}}+e^{iqx_{k-1}}\right)=0
\]
or
\[
\rho^{2}_{q}-1- 2\gamma\rho_{q} \left(e^{iqh}-2+e^{-iqh}\right)=0 .
\]
Since
\[
e^{iqh}-2+e^{-iqh}=\left(e^{iqh/2}-e^{-iqh/2}\right)^{2}=-4\sin^{2}
\frac{qh}{2},
\]
we obtain
\[
\rho^{2}_{q}+\left(8\gamma\sin^{2} \frac{qh}{2}\right)\rho_{q}-1=0.
\]
One of the roots of this quadratic equation, namely,
\[
\rho_{q}=-4\gamma\sin^{2} \frac{qh}{2}- \sqrt{16\gamma^2\sin^{4}
\frac{qh}{2}+1},
\]
is greater than $1$ in magnitude for any $q$ such that $\sin(qh/2)\neq 0$. Thus, Richardson's
method is (unconditionally) unstable.
\end{solution}


\subsubsection{Crank-Nicolson method} 

Let $\delta_{x}^{2}$ be a finite-difference operator defined by
\begin{equation}
\delta_{x}^{2}w_{kj}=w_{k+1, j}-2w_{kj}+w_{k-1,j}. \label{cc0}
\end{equation}
Consider the initial boundary value problem for
the heat equation
\begin{eqnarray}
&&\frac{\pr u}{\pr t}(x,t) = K\frac{\pr^{2} u}{\pr x^{2}}(x,t), \quad
0<x< L, \quad 0<0<T,   \label{c1} \\
&&u(0, t) = u(L, t)=0 \quad \hbox{for} \quad t>0,   \label{c2} \\
&&u(x, 0) = u_{0}(x).   \label{c3}
\end{eqnarray}
With the help of the above notation the (explicit) forward-difference method for Eq. (\ref{c1}) can be written as
\begin{equation}
\frac{w_{k,j+1}-w_{kj}}{\tau}-K\frac{\delta_{x}^{2}w_{kj}}{h^2}=0, \label{cc4}
\end{equation}
the (implicit) backward-difference method can be written as
\begin{equation}
\frac{w_{k,j+1}-w_{kj}}{\tau}-K\frac{\delta_{x}^{2}w_{k,j+1}}{h^2}=0. \label{cc5}
\end{equation}
A better method is obtained by averaging the forward-difference
method (\ref{cc4}) and the backward-difference method (\ref{cc5}).
Adding Eq. (\ref{cc4}) to Eq. (\ref{cc5}) with weights $1/2$, we
obtain the difference equation
\begin{equation}
\frac{w_{k,j+1}-w_{k,j}}{\tau}- \frac{K}{2h^{2}}\left(
\delta_{x}^{2}w_{kj}+\delta_{x}^{2}w_{k,j+1}\right)=0 \label{qq14}
\end{equation}
or, equivalently,
\begin{equation}
w_{k,j+1}-w_{k,j}-\frac{\gamma}{2}\left( w_{k+1,
j}-2w_{kj}+w_{k-1,j}+ w_{k+1, j+1}-2w_{k,j+1}+w_{k-1,j+1}\right)=0
\label{qq15a}
\end{equation}
where $\gamma=K\tau/h^2$.
This method is called the {\bf Crank-Nicolson method}. It can be
shown that the local truncation error of the  Crank-Nicolson method
is $O(\tau^{2}+h^{2})$ (see problem sheet).

   
In matrix form, Eq. (\ref{qq15a}) can be
written as
\begin{equation}
A{\bf w}^{(j+1)}=B{\bf w}^{(j)} \quad \hbox{for} \quad
j=0,1,2,\dots, \label{qq15}
\end{equation}
where
\[\begin{split}
A&=\left[
\begin{array}{cccccc}
1+\gamma &-\gamma/2 &0      &\dots  &\dots &0 \\
-\gamma/2 &1+\gamma &-\gamma/2 &\ddots  &     &\vdots \\
0      &-\gamma/2 &1+\gamma &-\gamma/2 &\ddots &\vdots \\
\vdots &\ddots &\ddots &\ddots &\ddots &0 \\
\vdots &       &\ddots &\ddots &\ddots &-\gamma/2 \\
0      &\dots  &\dots  &0      &-\gamma/2 &1+\gamma
\end{array}\right], \\
B&=\left[
\begin{array}{cccccc}
1-\gamma &\gamma/2 &0      &\dots  &\dots &0 \\
\gamma/2 &1-\gamma &\gamma/2 &\ddots  &     &\vdots \\
0      &\gamma/2 &1-\gamma &\gamma/2 &\ddots &\vdots \\
\vdots &\ddots &\ddots &\ddots &\ddots &0 \\
\vdots &       &\ddots &\ddots &\ddots &\gamma/2 \\
0      &\dots  &\dots  &0      &\gamma/2 &1-\gamma
\end{array}\right].
\end{split}\]
If we know ${\bf w}^{(j)}$, we can find ${\bf w}^{(j+1)}$ by solving the tridiagonal linear
system (\ref{qq15}) by the double-sweep method.

\vskip 0.5cm  
To investigate the stability of the Crank-Nicolson method, we employ
the Fourier method. If $z_{kj}=w_{kj}-\tilde{w}_{kj}$, then
\begin{equation}
z_{k,j+1}-z_{k,j}-\frac{\gamma}{2} \left(z_{k+1,
j}-2z_{kj}-z_{k-1,j}+ z_{k+1, j+1}-2z_{k,j+1}+z_{k-1,j+1}\right)=0
\label{qq16}
\end{equation}
for $k=1,2, \dots, N-1$ and $j=1,2, \dots$. We seek a particular
solution of (\ref{qq16}) in the form
\begin{equation}
z_{k,j}=\rho_{q}^{j}e^{iqx_{k}}, \quad q\in{\mathbb R}. \label{qq17}
\end{equation}
Then, the finite-difference method (\ref{qq15a}) is
stable, if
\[
\vert\rho_{q}\vert\leq 1
\]
for all $q$.

   
Substitution of (\ref{qq17}) into (\ref{qq16})
yields
\begin{multline}
0=e^{iqx_{k}}\left(\rho_{q}^{j+1}-\rho_{q}^{j}\right)\\
-\frac{\gamma}{2}\left[\rho_{q}^{j}
\left(e^{iqx_{k+1}}-2e^{iqx_{k}}+e^{iqx_{k-1}}\right)
+\rho_{q}^{j+1}\left(e^{iqx_{k+1}}-2e^{iqx_{k}}+e^{iqx_{k-1}}\right)
\right]
\end{multline}
or
\begin{multline*}
0=\rho_{q}-1- \frac{\gamma}{2} \left[e^{iqh}-2+e^{-iqh}+
\rho_{q}\left(e^{iqh}-2+e^{-iqh}\right) \right]\\
\Rightarrow
\quad 0=\rho_{q}-1-
\frac{\gamma}{2}(\rho_{q}+1)\left(e^{iqh}-2+e^{-iqh}\right).
\end{multline*}
Since
\[
e^{iqh}-2+e^{-iqh}=\left(e^{iqh/2}-e^{-iqh/2}\right)^{2}=-4\sin^{2}
\frac{qh}{2},
\]
we obtain
\[
\rho_{q}-1+(\rho_{q}+1) 2\gamma\sin^{2} \frac{qh}{2}=0 \quad
\Rightarrow \quad \rho_{q}=\frac{1-2\gamma\sin^{2}
\frac{qh}{2}}{1+2\gamma\sin^{2} \frac{qh}{2}}.
\]
To find out whether $\vert\rho_{q}\vert \leq 1$, we consider the function
$f(\alpha)=\frac{1-\alpha}{1+\alpha}$ for $\alpha\in[0,\infty)$. Since
$f'(\alpha)=-\frac{2}{(1+\alpha)^2} < 0$ for all $\alpha\in[0,\infty)$,
it is a decreasing function.
In addition to this, we have $f(0)=1$ and $\lim_{\alpha\to\infty}f(\alpha)=-1$.
We conclude that
$-1< f(\alpha) \leq 1$ for $\alpha\geq 0$. Therefore, $\vert\rho_{q}\vert \leq 1$
for all $q\in\mathbb{R}$, and
the Crank-Nicolson method is unconditionally stable.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\subsection{Consistency, stability, and convergence}

 
By definition, a finite-difference approximation to a differential equation
is {\bf consistent} with this differential equation if local truncation errors
tend to zero as the step size goes to zero, i.e.
\[
\max_{k,j}\vert \tau_{ki}(h,\tau)\vert \to 0 \ \ {\rm as} \ \ h,\tau \to 0.
\]
Finite difference equation for PDEs arising in initial value
problems may display a phenomenon which has no counterpart in
ordinary differential equations. Successive reduction of step
sizes in $x$ and $t$ may generate a finite difference solution
which is stable, but which may converge to the solution of a
different differential equation. For example, the Du Fort -
Frankel method for solving the heat equation
\begin{equation}
\frac{\partial u}{\partial t}- \frac{\partial^{2}u}{\partial
x^{2}}=0 \label{ww1}
\end{equation}
is given by
\[
\frac{w_{k,j+1}-w_{k,j-1}}{2\tau}- \frac{w_{k+1,
j}-w_{k,j-1}-w_{k,j+1}+w_{k-1,j}}{h^{2}}=0.
\]
It can be shown that this method is always stable and has
the local truncation error
\[
\tau_{kj}=O\left(\tau^2+h^2+\frac{\tau^2}{h^2}\right).
\]
Thus, the Du Fort-Frankel formula is consistent with Eq. (\ref{ww1}) only
if $\tau$ goes to zero faster than $h$. If they go to zero
at the same rate, so that $\tau/h=\beta$ is fixed, then this approximation
is consistent not with the diffusion equation but with the hyperbolic equation
\[
\frac{\partial u}{\partial t}- \frac{\partial^{2}u}{\partial
x^{2}}+ \beta^2\frac{\partial^{2}u}{\partial t^{2}}=0.
\]





    
By definition, a finite-difference method is said to be {\bf convergent}
if the total error of the method
\[
E=\max\limits_{k,j}\vert u_{kj}-w_{kj}\vert
\]
tends to zero as $h\to 0$ and $\tau\to 0$:
\[
E \to 0 \quad {\rm as} \quad h\to 0 \quad {\rm and} \quad \tau\to 0.
\]
We know that an approximation for the initial boundary value problem (\ref{c1})--(\ref{c3})
can be obtained using the forward-difference method
\begin{multline}
w_{k,j+1}=\left(1-2\gamma\right)w_{kj}+
\gamma\left( w_{k+1, j}+w_{k-1,j}\right)\\ {\rm for}
\quad k=1,2,\dots,N-1, 
j=0,1,\dots,M-1, \quad\quad \label{c4}
\end{multline}
\begin{align}
w_{0,j}&=w_{N,j}=0  \quad {\rm for}
\quad j=0,1,\dots,M \quad {\rm and} 
 \label{c5} \\w_{k,0}&=u_{0}(x_{k}) \quad {\rm for}
\quad k=1,2,\dots,N-1.\notag
\end{align}
Here $\gamma\equiv K\tau/h^{2}$.
Also, we know that the local truncation error for the method is
\begin{equation}
\tau_{kj}= \frac{u_{k,j+1}-u_{kj}}{\tau}-\frac{\gamma}{\tau}
\left(u_{k+1,j}-2u_{kj}+u_{k-1,j}\right)=O(\tau+h^{2}), \label{c6}
\end{equation}
so that the method is consistent,
and that the method is stable under the condition
\begin{equation}
\gamma \leq \frac{1}{2}. \label{c7}
\end{equation}

\begin{theorem}
If the stability condition (\ref{c7}) is satisfied,  the forward-difference method 
(\ref{c4}), (\ref{c5}) is convergent.
\end{theorem}

\begin{proof}
It follows from (\ref{c6}) that
\begin{equation}
u_{k,j+1}=(1-2\gamma)u_{kj}+\gamma
\left(u_{k+1,j}+u_{k-1,j}\right)+O(\tau^2+h^{2}\tau).  \label{c8}
\end{equation}
Let
\begin{equation}
z_{kj}=u_{kj}-w_{kj}. \label{c9}
\end{equation}
Then it follows from (\ref{c8}) and (\ref{c4}) that
\begin{equation}
z_{k,j+1}=(1-2\gamma)z_{kj}+\gamma
\left(z_{k+1,j}+z_{k-1,j}\right)+O(\tau^2+h^{2}\tau)  \label{c10}
\end{equation}
for $k=1,2,\dots,N-1$ and $j=0,1,\dots,M-1$.
In view of (\ref{c5}),
\begin{equation}
z_{0,j}=z_{N,j}=0,  \quad j=0,1,\dots,M \quad {\rm and} \quad
z_{k,0}=0, \quad k=1,2,\dots,N-1. \label{c11}
\end{equation}
If the stability condition (\ref{c7}) is satisfied, then
all coefficients on the right side of (\ref{c10}) are positive.

By definition of the big-$O$ notation, Eq. (\ref{c10}) implies that for
sufficiently small $\tau$ and $h$
\[
\left\vert z_{k,j+1}-\left[(1-2\gamma)z_{kj}+\gamma
\left(z_{k+1,j}+z_{k-1,j}\right)\right]\right\vert \leq A_{k,j+1} \, \vert\tau^2+h^{2}\tau\vert
\]
for some positive $A_{k,j+1}$. This and the inequality
\[
\vert a\vert = \vert b +(a-b)\vert\leq \vert b\vert + \vert a-b\vert
\]
(which is valid for any $a$ and $b$)
have the consequence that
\begin{equation}
\vert z_{k,j+1}\vert \leq\vert (1-2\gamma)z_{kj}+\gamma
\left(z_{k+1,j}+z_{k-1,j}\right)\vert +A_{k,j+1} \, \vert\tau^2+h^{2}\tau\vert  \label{c10a}
\end{equation}
for  $k=1,2,\dots,N-1$ and $j=0,1,\dots,M-1$. 
(Set $a=z_{k+1,j}$ and $b=\left[(1-2\gamma)z_{kj}+\gamma
\left(z_{k+1,j}+z_{k-1,j}\right)\right]$.)

\vskip 3mm  
Let ${\bf z}^{(j)}=(z_{1j}, z_{2j}, \dots, z_{N-1,j})^{T}$ and let
\[
\Vert {\bf z}^{(j)}\Vert =\Vert {\bf z}^{(j)}\Vert_{\infty}=
\max\limits_{1\leq k\leq N-1}\vert z_{kj}\vert .
\]
It follows from Eq. (\ref{c10}) that
\begin{eqnarray}
\Vert {\bf z}^{(j+1)} \Vert&=&\max\limits_{k}\vert z_{k,j+1} \vert\nonumber\\
&\leq&
\max\limits_{k}\left\{\left\vert
(1-2\gamma)z_{kj}+\gamma
\left(z_{k+1,j}+z_{k-1,j}\right)\right\vert + A_{j+1} \, \vert\tau^2+h^{2}\tau\vert\right\} \nonumber \\
&\leq&(1 -2\gamma)\max\limits_{k}\vert z_{kj}\vert+
\gamma\max\limits_{k}\vert z_{k+1,j}\vert +
\gamma\max\limits_{k}\vert z_{k-1,j}\vert\nonumber\\
&&\qquad\qquad
+A_{j+1} \, \vert\tau^2+h^{2}\tau\vert \nonumber \\
&\leq&\Vert {\bf z}^{(j)} \Vert+A_{j+1} \, \vert\tau^2+h^{2}\tau\vert. \nonumber
\end{eqnarray}
Here $A_{j+1}=\max\limits_{k}A_{k,j+1}$.
This inequality is valid for all $j=0,1,\dots,M-1$.
It follows that
\begin{eqnarray}
\Vert {\bf z}^{(j)} \Vert&\leq&
\Vert {\bf z}^{(j-1)} \Vert +A_{j} \, \vert\tau^2+h^{2}\tau\vert \nonumber \\
&\leq&\Vert {\bf z}^{(j-2)} \Vert +(A_{j}+A_{j-1}) \, \vert\tau^2+h^{2}\tau\vert \nonumber \\
&&\dots\quad \dots\quad \dots\quad \dots\quad \dots\quad \nonumber \\
&\leq&\Vert {\bf z}^{(0)} \Vert +(A_{j}+\dots +A_{1})\, \vert\tau^2+h^{2}\tau\vert \nonumber \\
&\leq&\Vert {\bf z}^{(0)} \Vert +Aj\tau \, \vert\tau+h^{2}\vert=\Vert {\bf z}^{(0)} \Vert +
A t_{j} \, \vert\tau+h^{2}\vert \nonumber
\end{eqnarray}
where
\[
A=\max_{j}A_{j}=\max_{k,j}A_{kj}.
\]
Since ${\bf z}_{0}=0$, we obtain
\begin{equation}
\Vert {\bf z}^{(j)} \Vert \leq A t_{j} \, \vert\tau+h^{2}\vert \leq A T \vert\tau+h^{2}\vert. \label{c13}
\end{equation}
Since this inequality is valid for all $j=0,1,\dots,M$, we obtain
\[
E= \max\limits_{k,j}\vert z_{kj} \vert=\max\limits_{j}\Vert {\bf z}^{(j)} \Vert\leq A T \vert\tau+h^{2}\vert \quad \Rightarrow \quad E\to 0 \quad {\rm as} \quad \tau,h\to 0.
\]
Thus, we have proved that the forward-difference method is convergent.
\end{proof}
Similar statements can be proved for other finite-difference methods.
 
Lax and Richtmyer\footnote{Lax, P.D., and Richtmyer, R.D. (1956). Survey of the stability of linear finite difference equations. Comm. Pure Appl. Math. 9, 267–293.
} and others studied the relation between consistency, stability,
and convergence of the approximations of linear initial value problems by
finite difference equations. The major result of that study is
\begin{theorem}[Lax equivalence theorem]
Given a well-posed initial boundary value problem and a finite
difference approximation to it that satisfies the consistency condition, then
stability is the necessary and sufficient condition for convergence.
\end{theorem}
