---
title: "Lab 4 Exercises"
output: html_notebook
---
```{r}
forwardDifference <- function(f=function(x) 0, 
                              u0=function(x) 2*sin(2*pi*x), 
                              K=1, L=1, N=30, T=0.1, M=200) {
    # set up space grid
    h <- L/N
    x <- h*(1:(N-1))
    
    # set up time grid
    tau <- T/M
    t <- tau*(0:M)
    
    # set up vectors with initial condition and inhomogenous term
    w <- u0(x)
    F <- f(x)
    
    # Set up evolution matrix
    gamma <- K*tau/(h^2)
    A <- diag(1-2*gamma, N-1)
    for (k in 1:(N-2)) {
        A[k,k+1] <- gamma
        A[k+1,k] <- gamma
    }
    
    Temperature <- matrix(0, N-1, M+1)  # Matrix to hold the solution
    Temperature[ , 1] <- w  # Initial value
    # Loop over time steps
    for (j in 1:M) {
        w <- A %*% w + tau * F
        Temperature[ , j+1] <- w
    }
    
    # Return a list consisting of time grid, x grid and solution
    return(list(x=x, t=t, w=Temperature))
}

library(plot3Drgl)
```

## Exercise 1
Use the function `forwardDifference()` to solve the equation
$$u_{t}(x,t) = \frac14 u_{xx}(x,t)-16\sin(8\pi x), \quad
0<x< 1, \quad 0<t<0.2$$
with initial condition
$$u(0, t) = u(1, t)=0, \quad u(x, 0) = \sin(\pi x).$$
Plot the solution.
```{r}
sol <- forwardDifference(f=function(x) -16*sin(8*pi*x), 
                  u0=function(x) sin(pi*x),
                  K=1/4, T=0.2)
persp3D(sol$x, sol$t, sol$w,
        xlab="x", ylab="t", zlab="w",
        ticktype="detailed", nticks=4)
plotrgl(smooth=TRUE, lighting = TRUE)
```


## Exercise 2
Determine the maximal error when you use a step-size of $h=1/120$. Keep on
doubling the number of steps in the $x$ directions. Does the error continue
to decrease as expected? What happens to the running time of the code?

For convenince I first encapsulate the code for calculating the maximal error
into a function so that we can then easily call it with different step sizes.
```{r}
maxError <- function(h) {
    # Determine number of steps and step size to guarantee stability
    N <- ceiling(1/h)
    K <- 1
    tau <- h^2/(2*K)
    T <- 0.2
    M <- ceiling(T/tau)
    
    # numerical solution
    numSol <- forwardDifference(f=function(x) {-25*sin(3*pi*x)},
                         u0=function(x) {-3/2*sin(2*pi*x)},
                         T=T, M=M, N=N)
    # exact solution
    x <- numSol$x
    t <- numSol$t
    xy <- mesh(x, t)
    u <- with(xy, -3/2*sin(2*pi*x)*exp(-4*pi^2*y)
                       -25/(9*pi^2)*sin(3*pi*x)*(1-exp(-9*pi^2*y)))
    
    return(max(abs(u - numSol$w)))
}
```

With h=1/120 we obtain a maximum error of
```{r}
maxError(1/120)
```
We observe that the error has again decreased by a factor of about 1/4.
So let's halve the step size again:

```{r}
maxError(1/240)
```

Another improvement by a factor of 1/4 but also a marked increase in runtime.


## Exercise 3
Similarly calculate the maximal error in the numerical solution from
exercise 1, again for a range of step-sizes.

The exact solution of the problem in exercise 1 is
$$
u(x,t) = \sin(\pi x)e^{-\pi^2 t}
-\frac{1}{4\pi^2}\sin(8\pi x)\left(1-e^{-64\pi^2 t}\right).
$$
We modify the function `maxError()` appropriately to reflect the
new problem.

```{r}
maxError <- function(h) {
    # Determine number of steps and step size to guarantee stability
    N <- ceiling(1/h)
    K <- 1
    tau <- h^2/(2*K)
    T <- 0.2
    M <- ceiling(T/tau)
    
    # numerical solution
    numSol <- forwardDifference(f=function(x) -16*sin(8*pi*x), 
                             u0=function(x) sin(pi*x),
                             K=K, T=T, M=M, N=N)
    # exact solution
    x <- numSol$x
    t <- numSol$t
    xy <- mesh(x, t)
    u <- with(xy, sin(pi*x)*exp(-pi^2*y)
                       -1/(4*pi^2)*sin(8*pi*x)*(1-exp(-64*pi^2*y)))
    
    return(max(abs(u - numSol$w)))
}
```

```{r}
maxError(1/30)
maxError(1/60)
maxError(1/120)
maxError(1/240)
```


## Exercise 4 (Challenge)
Write code that produces a plot of the maximal error against the number of 
steps $M$.

```{r}
h <- 1/(30*2^(0:3))
error <- sapply(h, maxError)
plot(h, error)
```

To see better that the error goes as $h^2$ it is a good idea to plot the log
of the error against the log of $h^2$. That should produce a straight line
with sloppe 2.
```{r}
plot(log(h), log(error))
abline(lm(log(error) ~ log(h)))
```
```{r}
summary(lm(log(error) ~ log(h)))
```

