---
title: "Numerical Methods for PDEs (Spring 2017), Lab 12"
author: "Gustav Delius, Richard Southwell"
output: html_notebook
---

# The method of lines

The "method of lines" for parabolic equations consists of at first only discretizing the $x$ direction but leaving the time continuous. This converts the PDEs into a system of ODEs. Then one uses a well-established method for the solution of systems of ODEs to get a numerical solution.

The main advantage of this approach is that there are very good packages for solving ODEs that implement powerful methods that would be a lot of work to code from scratch. In this lab we will use the deSolve package in R.
```{r}
#install.packages("deSolve")
library(deSolve)
```

We will illustrate the method with the system of PDEs from Lab 11:
$$
\begin{split}
\frac{\partial P}{\partial t} &= D\frac{\partial^2 P}{\partial x^2}
+h_- \frac{\partial}{\partial x}P\frac{\partial Z}{\partial x}
+\beta P(1-P)-Z\frac{P^2}{P^2+\nu^2},\\
\frac{\partial Z}{\partial t} &= D\frac{\partial^2 Z}{\partial x^2}
-h_+ \frac{\partial}{\partial x}Z\frac{\partial P}{\partial x}
+\gamma Z\left(\frac{P^2}{P^2+\nu^2}-w\right).
\end{split}
$$

We had seen that the forward-difference method has stability problems. Coding an explicit method for this set of non-linear equations would be a lot of work. So it is very tempting to use the method of lines.

We perform the discretization in the x direction as usual by choosing a step size $h$ and introducing points $x_k=k h$ for $k=0,\dots, N$.
```{r}
h <- 0.1
N <- 400
x <- (0:N)*h
```

We introduce $P_k(t)$ and $Z_k(t)$ as approximations of $P(x_k,t)$ and $Z(x_k,t)$, and by using the standard finite-difference expressions for the x derivatives we convert the system of PDEs into the following system of ODEs:
$$
\begin{split}
\frac{d P_{k}}{dt}=& \ 
\frac{D}{h^2}(P_{k+1}(t)-2P_{k}(t)+P_{k-1}(t))\\
&+\frac{h_-}{2h^2}\Big((P_{k+1}(t)-P_{k}(t))(Z_{k+1}(t)-Z_{k}(t))
-(P_{k}(t)+P_{k-1}(t))(Z_{k}(t)-Z_{k-1}(t))\Big)\\
&+\beta P_{k}(t)(1-P_{k}(t))-Z_{k}(t)\frac{P_{k}(t)^2}{P_{k}(t)^2+\nu^2}
\\
\frac{d Z_{k}}{dt}= 
&\frac{D}{h^2}(Z_{k+1}(t)-2Z_{k}(t)+Z_{k-1}(t))\\
&-\frac{h_+}{2h^2}\Big((Z_{k+1}(t)-Z_{k}(t))(P_{k+1}(t)-P_{k}(t))
-(Z_{k}(t)+Z_{k-1}(t))(P_{k}(t)-P_{k-1}(t))\Big)\\
&+\gamma Z_{k}(t)\left(\frac{P_{k}(t)^2}{P_{k}(t)^2+\nu^2}-w\right)
\end{split}
$$
for $k=0,\dots,N$. As discussed in lab 11, the Dirichlet boundary conditions give
$$
P_{-1}(t)=P_{1}(t),~~~~
P_{N+1}(t)=P_{N-1}(t),~~~~
Z_{-1}(t)=Z_{1}(t),~~~~
Z_{N+1}(t)=Z_{N-1}(t)
$$

We will use the function `ode()` from the deSolve package. This requires us to package the right-hand side of the ODE into a function. That function takes a vector $v=(P_0,P_1,\dots,P_N,Z_0,Z_1,\dots Z_N)$ and returns a vector of the corresponding time derivatives. We can reuse code from lab 11 to write this function.

```{r}
f <- function(t, y, parms) {
    N <- length(y)/2-1

    P <- y[1:(N+1)]
    Pp <- c(P[2:(N+1)], P[N])
    Pm <- c(P[2], P[1:N])
    
    Z <- y[(N+2):length(y)]
    Zp <- c(Z[2:(N+1)], Z[N-1])
    Zm <- c(Z[2], Z[1:N])
    
    Pd <- D/h^2 * (Pp - 2*P + Pm) + 
        hm/(2*h^2) * ((Pp+P)*(Zp-Z)-(P+Pm)*(Z-Zm)) +
        beta*P*(1-P)-Z*P^2/(P^2+nu^2)
    
    Zd <- D/h^2 * (Zp - 2*Z + Zm) - 
        hp/(2*h^2) * ((Zp+Z)*(Pp-P)-(Z+Zm)*(P-Pm)) +
        gamma*Z*(P^2/(P^2+nu^2) - w)
    
    list(c(Pd, Zd))
}
```
The reason we gave the function `f()` the arguments `t` and `parms` even though they are not used is that this is the form that `ode()` expects. This is also the reason why we wrapped the returned vector of time derivatives in a `list()`.

We also need to create a vector of times at which we want `ode()` to return values.
```{r}
t <- (0:100)*4
```

We choose parameter values
```{r}
D <- 0.04
nu <- 0.07
beta <- 1
gamma <- 0.01
w <- 0.004/gamma
hp <- 0
hm <- 0
```
and create a vector containing the initial values:
```{r}
P0 <- uniroot(function(P) P^2/(P^2+nu^2)-w, interval=c(0,1))$root
Z0 <- beta*P0*(1-P0)/w
Pn <- P0 + ifelse(x<1/4, 0.2*(cos(4*pi*x)+1), 0)
Zn <- Z0 + 0*x
y0 <- c(Pn, Zn)
```

We can now call the `ode()` function to obtain the solution.
```{r}
sol <- ode(y0, t, f, parms=NULL)
```

We extract the P and Z values from the returned array.
```{r}
P <- t(sol[, 2:(N+2)])
Z <- t(sol[, (N+3):(2*N+3)])
```

Now we can plot the prey abundance 
```{r}
library(plot3Drgl)
persp3D(x, t, P,
        ticktype="detailed", nticks=4)

plotrgl(lighting = TRUE, smooth = TRUE)
```
and the predator abundance
```{r}
persp3D(x, t, Z,
        ticktype="detailed", nticks=4)

plotrgl(lighting = TRUE, smooth = TRUE)
```

and the profile of both at an intermediate time:
```{r}
plot(x, P[, 51], type="l", ylab="")
lines(x, Z[, 51], lty="dotted")
legend("topleft", legend=c("P", "Z"), lty=c("solid", "dotted"))
```

This agrees with the corresponding plots from Lab 11.

Next we go to the parameter values that created stability issues in Lab 11.

```{r}
hp <- 0.9
N <- 520
x <- (0:N)/4
t <- (0:100)*3.9
Pn <- P0 + ifelse(x<1, 0.4*(cos(pi*x)+1), 0)
Zn <- Z0 + 0*x
y0 <- c(Pn, Zn)

sol <- ode(y0, t, f, parms=NULL)
P <- t(sol[, 2:(N+2)])
Z <- t(sol[, (N+3):(2*N+3)])
```



```{r}
persp3D(x, t, P,
        ticktype="detailed", nticks=4)
```

```{r}
persp3D(x, t, Z,
        ticktype="detailed", nticks=4)
plotrgl(lighting = TRUE, smooth = TRUE)
```

