---
title: "Numerical Methods for PDEs (Spring 2017), Lab 11"
author: "Gustav Delius, Richard Southwell"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this short lab is to reassure everyone that the even if you have
a complicated-looking system of parabolic PDEs, the forward-difference method
is straightforward to apply. It only has the drawback of requiring a very small
timestep. One may therefore like to later go to implicit methods, which are
much more clever and tricky and can give an amazing speed-up while at the same
time removing stability problems. However I recommend to always initially
set up the simple forward-difference method, if only to have something simple
to compare the more sophisticated methods against.

For this lab I chose the paper

Tsyganov, M.A., Brindley, J., Holden, A.V., and Biktashev, V.N. (2003). 
Quasisoliton Interaction of Pursuit-Evasion Waves in a Predator-Prey System. 
Phys. Rev. Lett. 91, 218102.

It describes the abundance $Z$ of a predator (you can think of zooplankton if
you like) and an abundance $P$ of prey (phytoplankton) by the set of
coupled equations
$$
\begin{split}
\frac{\partial P}{\partial t} &= D\frac{\partial^2 P}{\partial x^2}
+h_- \frac{\partial}{\partial x}P\frac{\partial Z}{\partial x}
+\beta P(1-P)-Z\frac{P^2}{P^2+\nu^2},\\
\frac{\partial Z}{\partial t} &= D\frac{\partial^2 Z}{\partial x^2}
-h_+ \frac{\partial}{\partial x}Z\frac{\partial P}{\partial x}
+\gamma\left(Z\frac{P^2}{P^2+\nu^2}-w\right).
\end{split}
$$
Here $D, h_-, h_+, \beta, \nu, \gamma$ are constant parameters. We choose
```{r}
D <- 0.04
nu <- 0.053
beta <- 0.43
gamma <- 0.05
w <- 0.34
hp <- 1
hm <- 1
```

We will impose Dirichlet boundary conditions.

If this was a proper project report we would now discuss the meaning of
these terms a bit, but for now we are only interested in implementing the
forward difference method for this somewhat complicated looking set of equations.

We introduce a grid of $x$ points
```{r}
h <- 0.02
N <- 50
x <- (1:(N-1))*h
```

We expect that we will need to choose a small time step $\tau$ for the explicit
forward-difference method. Presumably it needs to satisfy the inequality
$$
\tau \leq \frac{h^2}{2D}
$$
as in eq.(2.17) in the notes.

```{r}
tau <- h^2/(2*D)
```

Let us assume we want to integrate up to time $T=10$. 

```{r}
T <- 10
```

That requires

```{r}
M <- 10/tau
```

time steps. We do not want to save and plot so many results, so we split this
as $M = M_1 M_2$ and then only save $M_1 = 100$ results.

```{r}
M1 <- 100
M2 <- round(M/M1)
```

We create the vector of times at which we save the results

```{r}
t <- (0:M1)*M2*tau
```

We initialise two matrices to hold the results

```{r}
wp <- matrix(0, nrow=N-1, ncol=M1+1)
wz <- matrix(0, nrow=N-1, ncol=M1+1)
```

We will choose initial condition
$$
P(x,0) = 0.035 + \begin{cases}(\sin(4\pi x))/1000 & \text{ if } x<1/4\\
0& \text{ otherwise }\end{cases}, ~~~ Z(x,0) = 0.046.
$$
```{r}
Pn <- 0.035 + ifelse(x<1/4, (sin(4*pi*x))/1000, 0)
Zn <- rep(0.046, length(x))
```

Now we are ready to step through time and at each step update the result
according to the forward-difference formula and at every $M_2$-th step
save the result.
```{r}
for (j in 1:M1) {
    Pn <- wp[, j]
    Zn <- wz[, j]
    for (j2 in 1:M2) {
        
        P <- Pn
        Pp <- c(P[2:(N-1)], P[N-1])
        Pm <- c(P[1], P[1:(N-2)])
        
        Z <- Zn
        Zp <- c(Z[2:(N-1)], Z[N-1])
        Zm <- c(Z[1], Z[1:(N-2)])
        
        Pn <- P + tau * (
            D/h^2 * (Pp - 2*P + Pm) + 
            hm/(2*h^2) * ((Pp+P)*(Zp-Z)-(P+Pm)*(Z-Zm)) +
            beta*P*(1-P)-Z*P^2/(P^2+nu^2)
            )
        
        Zn <- Z + tau * (
            D/h^2 * (Zp - 2*Z + Zm) - 
            hm/(2*h^2) * ((Zp+Z)*(Pp-P)-(Z+Zm)*(P-Pm)) +
            gamma*Z*(P^2/(P^2+nu^2) - w)
            )
    }
    wp[, j+1] <- Pn
    wz[, j+1] <- Zn
}
```

Finally we make a plot of the result. Here is the predator abundance:
```{r}
library(plot3Drgl)
persp3D(x, t, wz,
        xlab="x", ylab="t", zlab="Z",
        ticktype="detailed", nticks=4)

plotrgl(lighting = TRUE)
```

Here is the prey abundance:
```{r}
persp3D(x, t, wp,
        xlab="x", ylab="t", zlab="P",
        ticktype="detailed", nticks=4)

plotrgl(lighting = TRUE)
```




