---
title: "Numerical Methods for PDEs (Spring 2017), Practical 4"
author: "Richard Southwell, Gustav Delius"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

# Encapsulating code in a function

In the last lab you wrote code to use the forward difference method to
numerically solve the heat equation
$$ u_{t}(x,t) = K u_{xx}(x,t)+f(x),$$
for $0<x< L, 0<t<T$, subject to the boundary conditions
$$ u(0, t) = u(L, t)=0, $$
and initial condition
$$ u(x, 0) = u_0(x). $$
To make it easy to re-use that code to obtain solutions with different
parameters and different choices for the initial condition $u_0(x)$ and
the inhomogenous term $f(x)$ we encapsulate that code inside a function.
The parameters that we may want to change we make into arguments for the
function. 

```{r}
forwardDifference <- function(f=function(x) 0, 
                              u0=function(x) 2*sin(2*pi*x), 
                              K=1, L=1, N=30, T=0.1, M=200) {
    # set up space grid
    h <- L/N
    x <- h*(1:(N-1))
    
    # set up time grid
    tau <- T/M
    t <- tau*(0:M)
    
    # set up vectors with initial condition and inhomogenous term
    w <- u0(x)
    F <- f(x)
    
    # Set up evolution matrix
    gamma <- K*tau/(h^2)
    A <- diag(1-2*gamma, N-1)
    for (k in 1:(N-2)) {
        A[k,k+1] <- gamma
        A[k+1,k] <- gamma
    }
    
    Temperature <- matrix(0, N-1, M+1)  # Matrix to hold the solution
    Temperature[ , 1] <- w  # Initial value
    # Loop over time steps
    for (j in 1:M) {
        w <- A %*% w + tau * F
        Temperature[ , j+1] <- w
    }
    
    # Return a list consisting of time grid, x grid and solution
    return(list(x=x, t=t, w=Temperature))
}
```

There are two things to note about the function arguments:
1) R allows the arguments to functions to themselves be functions. In the
above function the arguments f and u0 are functions.
2) R allows you to specify default values for some arguments. Then you can
call the function without having to specify those arguments where you just
want to use the default values

Let us use this function to reproduce the first example from Lab 3

```{r}
sol <- forwardDifference()
```

Note that we did not have to specify any of the arguments because we were
happy with the default values. 

Next we plot the solution:

```{r}
library("plot3Drgl") 
persp3D(sol$x, sol$t, sol$w,
        xlab="x", ylab="t", zlab="w",
        ticktype="detailed", nticks=4)
plotrgl(smooth=TRUE, lighting = TRUE)
```

Next let's reproduce the the example from the last exercise from Lab 3. 

```{r}
sol <- forwardDifference(f=function(x) {-25*sin(3*pi*x)},
                         u0=function(x) {-1.5*sin(2*pi*x)},
                         T=0.2, M=600
)
persp3D(sol$x, sol$t, sol$w,
        xlab="x", ylab="t", zlab="w",
        ticktype="detailed", nticks=4)
```

## Exercise 1
Use the function `forwardDifference()` to solve the equation
$$u_{t}(x,t) = \frac14 u_{xx}(x,t)-16\sin(8\pi x), \quad
0<x< 1, \quad 0<t<0.2$$
with initial condition
$$u(0, t) = u(1, t)=0, \quad u(x, 0) = \sin(\pi x).$$
Plot the solution.



# Studying the error

Consider again the example from the exercise in Lab 3:
$$u_{t}(x,t) = u_{xx}(x,t)-25 \sin(3\pi x), \quad
0<x< 1, \quad 0<t<0.2,$$
with initial condition
$$u(0, t) = u(1, t)=0, \quad u(x, 0) = -\frac32\sin(2\pi x).$$

The exact solution is
$$u(x,t) = -\frac32 \sin(2\pi x)e^{-4\pi^2 t}
-\frac{25}{9\pi^2}\sin(3\pi x)\left(1-e^{-9\pi^2 t}\right).$$

We will now compare the numerical solution to the exact solution. The numerical solution is

```{r}
numSol <- forwardDifference(f=function(x) {-25*sin(3*pi*x)},
                         u0=function(x) {-3/2*sin(2*pi*x)},
                         T=0.2, M=360
)
```

To evaluate the exact solution at the grid points we first construct the grid (mesh)

```{r}
x <- numSol$x
t <- numSol$t
xy <- mesh(x, t)
```

and then create the matrix of values at the grid points

```{r}
u <- with(xy, -3/2*sin(2*pi*x)*exp(-4*pi^2*y)
                       -25/(9*pi^2)*sin(3*pi*x)*(1-exp(-9*pi^2*y)))
```

When we plot the exact solution it looks just like the numerical solutions we
had plotted already above.

```{r}
persp3D(x, t, u, zlab="u", ticktype="detailed", nticks=4) 
```

But when we plot the difference between the numerical solution and the exact
solution we see that they were in fact not quite equal. 

```{r}
persp3D(x, t, u - numSol$w, zlab="u - w", ticktype="detailed", nticks=4)
```

There is quite a large error. The maximal error is

```{r}
max(abs(u - numSol$w))
```

This relatively large error is due to the large stepsize we used in the
$x$ direction. When we double the number of steps from $n=60$ to $N=60$ we
should see a reduction in the error. Unfortunately we will also need to
reduce the step size in the $t$ direction, so that the stability condition
$$ \tau < \frac{h^2}{2K}$$
is satisfied. In our case $h=L/N=1/60$ and $K=1$, so we need $\tau=T/M<1/7200$
or equivalently $M>7200 T$. With $T=0.2$ this gives $M > 1440$. 
Just out of fun let us check that we really can't get away with fewer time steps.
Let's see what happens if we use 1400 time steps instead of the required 1440.
```{r}
numSol <- forwardDifference(f=function(x) {-25*sin(3*pi*x)},
                         u0=function(x) {-3/2*sin(2*pi*x)},
                         T=0.2, M=1400, N=60
)
persp3D(numSol$x, numSol$t, numSol$w, 
        xlab="x", ylab="t", zlab="w", 
        ticktype="detailed", nticks=4) 
```

As expected, we get nonsense. So let's redo the calculation with 1440 time steps.

```{r}
numSol <- forwardDifference(f=function(x) {-25*sin(3*pi*x)},
                         u0=function(x) {-3/2*sin(2*pi*x)},
                         T=0.2, M=1440, N=60
)
persp3D(numSol$x, numSol$t, numSol$w, 
        xlab="x", ylab="t", zlab="w", 
        ticktype="detailed", nticks=4) 
```

Notice that now we have approximations $w$ at many more gridpoints. The
resolution in the $t$ direction is much larger than we need to produce a nice
plot. This unnecessarily slows down the production of the plot, so we will
work only with a subset of all the $t$ values. Out of the
1441 values we will plot only every 4th. Luckily selecting a subset of 
entries of a vector or a matrix is very easy in R. We first introduce the
technique in a simple example.
```{r}
v <- (1:10)^2
v
```
Here is how we would select the third and sixth element from the vector `v`:
```{r}
v[c(3, 6)]
```
To select only all elements with odd indices we can do
```{r}
select <- seq(1, 10, by=2)
v[select]
```
We can now apply this technique to extracting every 4th element in the $t$
direction.
```{r}
tSelect <- seq(1, 1441, by=4)
t <- numSol$t[tSelect]
w <- numSol$w[, tSelect]
```

We calculate the exact solution at the same gridpoints

```{r}
x <- numSol$x
xy <- mesh(x, t)
u <- with(xy, -3/2*sin(2*pi*x)*exp(-4*pi^2*y)
                       -25/(9*pi^2)*sin(3*pi*x)*(1-exp(-9*pi^2*y)))
```

Now we can plot the difference between the exact solution and the numerical
solution

```{r}
persp3D(x, t, u - w, zlab="u - w", ticktype="detailed", nticks=4)
```
The error has indeed been reduced by a factor of about $4$ by decreasing the 
step-size $h$ by a factor of $2$. For example the maximal error is now

```{r}
max(abs(u - w))
```

## Exercise 2
Determine the maximal error when you use a step-size of $h=1/120$. Keep on
doubling the number of steps in the $x$ directions. Does the error continue
to decrease as expected? What happens to the running time of the code?

## Exercise 3
Similarly calculate the maximal error in the numerical solution from
exercise 1, again for a range of step-sizes.

## Exercise 4 (Challenge)
Write code that produces a plot of the maximal error against the number of
steps $M$.

