---
title: "Numerical Methods for PDEs (Spring 2017), Lab 8"
author: "Richard Southwell, Gustav Delius"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

# Non-linear heat equation

In this lab you will explore the backward difference method for the
non-linear heat equation

$$
\frac{\partial u}{\partial t} =\frac{\partial }{\partial x}\left(K(x,t,u)
\frac{\partial u}{\partial x}\right)+f(x,t,u),  \quad 0<x< L, \quad t>0,
$$
subject to the initial and boundary conditions
$$
u(x, 0) = u_{0}(x), \qquad u(0,t)=0, \quad u(L,t)=0.
$$

## Semi-implicit method

We will extend the function `backwardDifference()` from Lab 5 to allow
$K$ and $f$ to be functions of $x, t$ and $u$. Initially we will use
the method from eq.(2.106) in the lecture notes, which evaluates
$K$ and $f$ at the $w$ from the previous time step in order to avoid
having to solve a non-linear equation at each time step. Such a method is
sometimes called "semi-implicit". 

```{r}
backwardDifference2 <- function(u0, K, f, L=1, N=30, T=1, M=30) {
    # set up space grid
    h <- L/N
    x <- h*(1:(N-1))
    xFull <- h*(0:N)
    
    # set up time grid
    tau <- T/M
    t <- tau*(0:M)
    
    gamma <- tau/h^2
    
    w <- matrix(0, N-1, M+1)  # Matrix to hold the solution
    w[ , 1] <- u0(x)  # Initial value
    # Loop over time steps
    for (j in 1:M) {
        Kv <- K(xFull, t[j], c(0, w[, j], 0)) 
        chi <- (Kv[1:N] + Kv[2:(N+1)])/2
        a <- 1 + gamma*(chi[2:N]+chi[1:(N-1)])
        b <- -gamma*chi[1:(N-1)]
        w[, j+1] <- doublesweep(b[1:(N-1)], c(b[2:(N-1)],0), -a, 
                                w[, j]+tau*f(x, t[j], w[, j]), 0, 0)
    }
    
    # Return a list consisting of time grid, x grid and solution
    return(list(x=x, t=t, w=w))
}
```

We will use the implementation of the doublesweep method from Lab 5:

```{r}
doublesweep <- function(A, B, C, F, a, b) {
    # Solves the equation 
    # A[i]*v[i-1] - C[i]*v[i] + B[i]*v[i+1] = F[i]
    # for v[i], i = 1,...,N-1, with boundary conditions
    # v[0]=a and v[N]=b
    
    # Check the lengths of the vectors
    N <- length(C) + 1
    if ((length(B) != N-1) || (length(A) != N-1) || (length(F) != N-1)) {
        stop("The lengths of the vector arguments need to be equal")
    }

    alpha <- rep(0, N)
    beta <- rep(0, N)
    beta[1] <- a
    
    #sweep up to get alphas and betas
    for (i in 1:(N-1)) {
        alpha[i+1] <- B[i] / (C[i]-alpha[i]*A[i])
        beta[i+1] <- (beta[i]*A[i] - F[i]) / (C[i] - alpha[i]*A[i])
    }
    
    v <- rep(0, N-1 )
    v[N-1] <- alpha[N]*b + beta[N]
    
    #sweep down to find v's
    for (i in (N-1):2) {
        v[i-1] <- alpha[i]*v[i] + beta[i]    
    }
    
    return(v)
}
```
Let us look at the solution of the equation
$$
\frac{\partial u}{\partial t} =\frac{\partial }{\partial x}\left(\frac12 u^2
\frac{\partial u}{\partial x}\right)-u,  \quad 0<x< L, \quad t>0,
$$
with initial condition 
$$
u(x,0) = \max(0, 1-10|x-0.5|)
$$

```{r}
sol <- backwardDifference2(u0=function(x) pmax(0, 1-10*abs(x-0.5)), 
                           K=function(x, t, u) u^2/2,
                           f=function(x, t, u) -u)
library("plot3Drgl") 
persp3D(sol$x, sol$t, sol$w,
        xlab="x", ylab="t", zlab="w",
        ticktype="detailed", nticks=4)
```

### Exercise 1
Solve the equation 
$$
\frac{\partial u}{\partial t} =\frac{\partial }{\partial x}\left(x\,t\,u^2
\frac{\partial u}{\partial x}\right)+8x\sin(8\pi xt),  \quad 0<x< 1, \quad 0<t<1,
$$
with the same initial condition as above.


## Method of successive approximations

We now add another loop to our function `backwardDifference2()` that
implements the method of successive approximations, see eq.(2.107) in
the lecture notes.

We introduce two extra arguments to the function: `max_iterations` is the
maximum number of iterations for the successive approximation method and
`tolerance` is the amount of error that is acceptable. So the iterations
are stopped as soon as the result varies by less than `tolerance`.

```{r}
backwardDifference2 <- function(u0=function(x) pmax(0,1-10*abs(x-0.5)), 
                                K=function(x, t, u) u^2/2,
                                f=function(x, t, u) -u^2,
                                L=1, N=30, T=1, M=30,
                                max_iteration=20, tolerance=0.000001) {
    # set up space grid
    h <- L/N
    x <- h*(1:(N-1))
    xFull <- h*(0:N)
    
    # set up time grid
    tau <- T/M
    t <- tau*(0:M)
    
    gamma <- tau/h^2
    
    w <- matrix(0, N-1, M+1)  # Matrix to hold the solution
    iterations <- rep(0, M)  # to hold number of iterations needed at each step
    precision <- rep(0, M)  # to hold precision achieved
    
    w[ , 1] <- u0(x)  # Initial value
    # Loop over time steps
    for (j in 1:M) {
        wn <- w[, j]  # initial guess for w_{j+1}
        for (s in 1:max_iteration) {
            ws <- wn
            Kv <- K(xFull, t[j], c(0, ws, 0))
            chi <- (Kv[1:N] + Kv[2:(N+1)])/2
            a <- 1 + gamma*(chi[2:N]+chi[1:(N-1)])
            b <- -gamma*chi[1:(N-1)]
            # update guess for w_{j+1}
            wn <- doublesweep(b[1:(N-1)], c(b[2:(N-1)],0), -a, 
                              w[, j]+tau*f(x, t[j], ws), 0, 0)
            # break if tolerance limit is reached
            if (max(abs(wn-ws)) < tolerance) {
                break
            }
        }
        iterations[j] <- s
        precision[j] <- max(abs(wn-ws))
        w[, j+1] <- wn
    }
    
    # Return a list consisting of time grid, x grid and solution
    return(list(x=x, t=t, w=w, iterations=iterations, precision=precision))
}
```

Note that we have also added two more items to the returned list:
`iterations` lists the number of iterations actually performed at each time
step and `precision` is the amount by which the approximation changed during
the final iteration at each time step.

Let us use this modified function on our simple example

```{r}
sol <- backwardDifference2(u0=function(x) pmax(0, 1-10*abs(x-0.5)), 
                           K=function(x, t, u) u^2/2,
                           f=function(x, t, u) -u)

persp3D(sol$x, sol$t, sol$w,
        xlab="x", ylab="t", zlab="w",
        ticktype="detailed", nticks=4)
```

Let us look at how many iterations are used in the method of successive
approximations in order to reach the tolerance limit of 0.000001
```{r}
sol$iterations
```

Not surprisingly we wee that more iterations are needed in the initial time steps during which the solution changes a lot between steps. 

```{r}
sol$precision
```

If we had allowed only a single iteration the precision would have been poor:

```{r}
sol <- backwardDifference2(u0=function(x) pmax(0, 1-10*abs(x-0.5)), 
                           K=function(x, t, u) u^2/2,
                           f=function(x, t, u) -u,
                           max_iteration = 1)
sol$precision
```
```{r}
sol <- backwardDifference2(u0=function(x) pmax(0, 1-10*abs(x-0.5)), 
                           K=function(x, t, u) u^2/2,
                           f=function(x, t, u) -u,
                           max_iteration = 2)
sol$precision
```

### Exercise 2
How many iterations are required for the example from exercise 1 with
a tolerance of $10^{-6}$? What precision is achieved with just 4 iterations?

