---
title: "Nora's system solved with the method of lines"
author: "Gustav Delius"
output: html_notebook
---

# The method of lines

The "method of lines" for parabolic equations consists of at first only discretizing the $x$ direction but leaving the time continuous. This converts the PDEs into a system of ODEs. Then one uses a well-established method for the solution of systems of ODEs to get a numerical solution.

The main advantage of this approach is that there are very good packages for solving ODEs that implement powerful methods that would be a lot of work to code from scratch. In this lab we will use the deSolve package in R.
```{r}
#install.packages("deSolve")
library(deSolve)
library(plot3Drgl)
```

We perform the discretization in the x direction as usual by choosing a step size $h$ and introducing points $x_k=k h$ for $k=0,\dots, N$.
```{r}
h <- 0.05
N <- 120
x <- (0:N)*h
```
for $k=0,\dots,N$. As discussed in lab 11, the Dirichlet boundary conditions give
$$
P_{-1}(t)=P_{1}(t),~~~~
P_{N+1}(t)=P_{N-1}(t),
$$
and similarly for the other functions. However it seems Nora is using slightly different boundary conditions.

We will use the function `ode()` from the deSolve package. This requires us to package the right-hand side of the ODE into a function. That function takes a vector $v=(P_0,P_1,\dots,P_N,Q_0,Q_1,\dots Q_N,\dots)$ and returns a vector of the corresponding time derivatives. We can combine the code from lab 11 with the code written by Nora to define this function:

```{r}
f <- function(t, y, parms) {
    N <- length(y)/6-1
    
    P <- y[1:(N+1)]
    Pp <- c(P[2:(N+1)], P[N])       #For P_k+1 entries
    Pm <- c(P[2], P[1:N])           #For P_k-1 entries
    
    Q <- y[(N+2):(2*N+2)]
    Qp <- c(Q[2:(N+1)], Q[N-1])     #For Q_k+1 entries
    Qm <- c(Q[2], Q[1:N])           #For Q_k-1 entries
    
    A <- y[(2*N+3):(3*N+3)]
    Ap <- c(A[2:(N+1)], A[N])       #For A_k+1 entries
    Am <- c(A[2], A[1:N])           #For A_k-1 entries
    
    C <- y[(3*N+4):(4*N+4)]
    Cp <- c(C[2:(N+1)], C[N-1])     #For C_k+1 entries
    Cm <- c(C[2], C[1:N])           #For C_k-1 entries
    
    H <- y[(4*N+5):(5*N+5)]
    Hp <- c(H[2:(N+1)], H[N])       #For H_k+1 entries
    Hm <- c(H[2], H[1:N])           #For H_k-1 entries
    
    R <- y[(5*N+6):(6*N+6)]
    Rp <- c(R[2:(N+1)], R[N-1])     #For Q_k+1 entries
    Rm <- c(R[2], R[1:N])           #For Q_k-1 entries
    
    Pn <- (-k/(2*h^2)) * ((Pp+P)*(Cp-C) - (P+Pm)*(C-Cm)) - 
        (v*A0/(2*h^2)) * ((Pp+P)*(Ap-A) - (P+Pm)*(A-Am)) + 
        #Alternate way to approximate the d/dx(AP*dA/dx) term
        #v*((A*P/h^2)*(Ap-2*A+Am) + ((Ap-Am)/2*h)*((A/2*h)*(Pp-Pm) + (P/2*h)*(Ap-Am))) + 
        (v/(2*h^2)) * ((Ap+A)*(Pp+P)*(Ap-A) - (A+Am)*(P+Pm)*(A-Am)) + 
        (mu/(h^2)) * (Pp - 2*P + Pm) - 
        r1*(1+sigma*A)*P - 
        w1*P + 
        ff
    
    Qn <- r1*(1 + sigma*A)*P - w2*Q - beta*Q*R
    
    An <- (b1/(h^2))*(Ap - 2*A + Am) + a1*Q - d1*A
    
    Cn <- (b2/(h^2))*(Cp - 2*C + Cm) + a2*r3*H*R - d2*C
    
    Hn <- r1*(1 + sigma*A)*P - r4*H*R
    
    Rn <- r2*(R0 - R)*R - r3*H*R
    
    list(c(Pn, Qn, An, Cn, Hn, Rn))
}
```
The reason we gave the function `f()` the arguments `t` and `parms` even though they are not used is that this is the form that `ode()` expects. This is also the reason why we wrapped the returned vector of time derivatives in a `list()`.

We also need to create a vector of times at which we want `ode()` to return values.
```{r}
t <- (0:100)*0.005
```

We choose parameter values
```{r}
#The values on the sides are the original values from the paper
k <- 10          #10
mu <- 2         #2
sigma <- 5      #5
a1 <- 24        #2400
d1 <- 20        #500
b2 <- 50         #50
R0 <- 1         #1
r2 <- 0.3       #0.1
r4 <- 0.2       #0.1
w2 <- 5         #0
v <- 50         #50
A0 <- 1         #1
gamma <- 0.2    #0.2
b1 <- 50         #50
a2 <- 5         #50
d2 <- 500       #500
r1 <- 0.513     #0.0513
r3 <- 10        #25
w1 <- 4     #0.0103
beta <- 20      #20
ff <- 1          #1
```

and create a vector containing the initial values:

```{r}
#Initial conditions
Pn <- 20*(exp(-(x-1)^2) + exp(-(x+1)^2))
Qn <- 1+0*x
An <- 0.2*exp(-x^2)
Cn <- 0.2*exp(-x^2)
Hn <- 1+0*x
Rn <- -0.5*exp(-x^2)+1

y0 <- c(Pn, Qn, An, Cn, Hn, Rn)
```

We can now call the `ode()` function to obtain the solution.
```{r}
sol <- ode(y0, t, f, parms=NULL)
```

We extract the values from the returned array.
```{r}
wp <- t(sol[, 2:(N+2)])
wq <- t(sol[, (N+3):(2*N+3)])
wa <- t(sol[, (2*N+4):(3*N+4)])
wc <- t(sol[, (3*N+5):(4*N+5)])
wh <- t(sol[, (4*N+6):(5*N+6)])
wr <- t(sol[, (5*N+7):(6*N+7)])
```

Now we can plot the abundances
```{r}
library(plot3Drgl)
persp3D(x, t, wp, zlab="P",
        ticktype="detailed", nticks=4)
```
```{r}
persp3D(x, t, wq, zlab="Q",
        ticktype="detailed", nticks=4)
```
```{r}
persp3D(x, t, wa, zlab="A",
        ticktype="detailed", nticks=4)
```
```{r}
persp3D(x, t, wc, zlab="C",
        ticktype="detailed", nticks=4)
```
```{r}
persp3D(x, t, wh, zlab="H",
        ticktype="detailed", nticks=4)
```
```{r}
persp3D(x, t, wr, zlab="R",
        ticktype="detailed", nticks=4)
```
and the profile of all at the final time:
```{r}
tn <- 101
plot(x, wp[,tn]/max(wp[,tn]), type='l', ylab='x', ylim=c(0, 1),
     main=paste('Time=', t[tn]))
lines(x, wq[,tn]/max(wq[,tn]), lty='dashed')
lines(x, wa[,tn]/max(wa[,tn]), lty='dotted')
lines(x, wc[,tn]/max(wc[,tn]), lty='dotdash')
lines(x, wh[,tn]/max(wh[,tn]), lty='longdash')
lines(x, wr[,tn]/max(wr[,tn]), lty='twodash')
legend('topright',legend=c('P','Q','A','C','H','R'),
       lty=c('solid','dashed','dotted','dotdash','longdash','twodash'))

```

